---
title: "SoccerFouls"
---

Do fouls matter in women's college soccer? In this assignment, you'll be working with a dataset that contains information about soccer matches from last season. You'll load the data, perform some basic statistical analysis, and create a visualization to explore the relationship between fouls and the goal differential (and thus the outcome). Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load the soccer dataset from https://raw.githubusercontent.com/dwillis/NCAAWomensSoccerData/refs/heads/main/data/ncaa_womens_soccer_matchstats_2024.csv

```{r}
matches <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensSoccerData/refs/heads/main/data/ncaa_womens_soccer_matchstats_2024.csv")
```


### Task 3: Create a new column called `goal_diff` that calculates the difference between the two teams' goals.

```{r}
matches <- matches |> mutate(goal_diff = team_score - opponent_score)
```



### Task 4: Use `summarize()` and the `cor()` function to calculate the correlation between fouls and goal differential, using the pearson method.

```{r}
matches |> summarize(correlation = cor(fouls, goal_diff, method="pearson"))
```



### Task 5: Create a scatterplot of fouls and goal differential, using `geom_point()`

```{r}

matches |> 
  ggplot() +
  geom_point(aes(x=goal_diff, y=fouls)) +
  geom_smooth(aes(x=goal_diff, y=fouls), method="lm")

```



### Task 6: Describe the results of both the correlation calculation and the scatterplot.


When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
