---
title: "Hypothesis Testing: Do Goggles Help Jeffrey Swim Faster?"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

In this analysis, we'll investigate whether wearing goggles helps Jeffrey swim faster. We'll compare his swim times with goggles to his baseline performance without goggles using a one-sample t-test.

## Step 1: Load Required Libraries

First, we need to load the tidyverse package, which includes ggplot2 for visualization and dplyr for data manipulation.

```{r}
# Load required libraries
library(tidyverse)
```

## Step 2: Create the Dataset

We'll create a dataset with Jeffrey's 15 swim times when wearing goggles.

```{r}
# Create the dataset with Jeffrey's 15 swim times with goggles
jeffrey_times <- c(
  16.2, 15.8, 16.1, 15.7, 16.3, 
  15.9, 16.0, 15.6, 16.4, 15.8, 
  16.1, 15.7, 15.9, 16.2, 16.0
)
```

## Step 3: Set Up Hypothesis Test Parameters

We'll define our baseline statistics (Jeffrey's performance without goggles) and significance level.

```{r}
# Set up our hypothesis test parameters
baseline_mean <- 16.43  # Jeffrey's mean time without goggles
baseline_sd <- 0.8      # Standard deviation without goggles
alpha <- 0.05           # Significance level
```

## Step 4: Organize Data into a Tibble

We'll create a structured tibble for easier analysis.

```{r}
# Create a tibble with the swim data
swim_data <- tibble(
  trial = 1:15,
  time_with_goggles = jeffrey_times
)

# Display the first few rows
head(swim_data)
```

## Step 5: Calculate Summary Statistics

Now we'll compute the key statistics needed for our hypothesis test.

```{r}
# Calculate summary statistics
swim_summary <- swim_data %>%
  summarize(
    sample_size = n(),
    sample_mean = mean(time_with_goggles),
    sample_sd = sd(time_with_goggles),
    se = sample_sd / sqrt(sample_size),
    t_stat = (sample_mean - baseline_mean) / se,
    degrees_freedom = sample_size - 1,
    p_value = pt(t_stat, degrees_freedom) # one-tailed test since we're testing if time is faster (lower)
  )

# Display the summary statistics
swim_summary
```

## Step 6: Visualize the Hypothesis Test

This visualization shows the null hypothesis distribution (blue curve) and where our sample mean falls (red dashed line).

```{r}
# Create a visualization of the hypothesis test
ggplot() +
  # Draw the null hypothesis distribution
  stat_function(
    fun = function(x) dnorm(x, mean = baseline_mean, sd = baseline_sd / sqrt(15)),
    geom = "line",
    color = "blue",
    size = 1
  ) +
  # Add a vertical line for the sample mean
  geom_vline(
    xintercept = swim_summary$sample_mean,
    color = "red",
    linetype = "dashed",
    size = 1
  ) +
  # Add annotation for p-value
  annotate(
    "text",
    x = swim_summary$sample_mean - 0.2,
    y = 0.4,
    label = paste("p-value =", round(swim_summary$p_value, 4)),
    color = "black"
  ) +
  # Add labels and title
  labs(
    title = "Hypothesis Test: Do Goggles Help Jeffrey Swim Faster?",
    subtitle = paste0("H₀: μ = ", baseline_mean, " seconds vs. H₁: μ < ", baseline_mean, " seconds"),
    x = "Swim Time (seconds)",
    y = "Probability Density"
  ) +
  # Set x-axis limits for better visualization
  xlim(15, 17.5) +
  theme_minimal()
```

## Step 7: Verify with R's Built-in t-test Function

We'll double-check our observations using R's built-in t.test function.

```{r}
# Perform the t-test using R's built-in function (for verification)
t_test_result <- t.test(
  swim_data$time_with_goggles,
  mu = baseline_mean,
  alternative = "less"  # one-sided test (faster means less time)
)

# Display the t-test results
t_test_result
```

## Step 8: Summarize Results in a Table

Finally, we'll create a comprehensive table summarizing all our findings.

```{r}
# Create a results table
results_table <- tibble(
  Metric = c(
    "Baseline Mean (without goggles)",
    "Baseline Standard Deviation",
    "Sample Mean (with goggles)",
    "Sample Standard Deviation",
    "Sample Size",
    "Standard Error",
    "t-statistic",
    "Degrees of Freedom",
    "p-value",
    "Significant at α = 0.05?",
    "Conclusion"
  ),
  Value = c(
    baseline_mean,
    baseline_sd,
    round(swim_summary$sample_mean, 2),
    round(swim_summary$sample_sd, 2),
    swim_summary$sample_size,
    round(swim_summary$se, 3),
    round(swim_summary$t_stat, 3),
    swim_summary$degrees_freedom,
    round(swim_summary$p_value, 4),
    ifelse(swim_summary$p_value < alpha, "Yes", "No"),
    ifelse(swim_summary$p_value < alpha, 
           "Reject H₀: Goggles helped Jeffrey swim faster", 
           "Fail to reject H₀: No evidence goggles helped")
  )
)

# Display the results table in a formatted way
knitr::kable(results_table, caption = "Summary of Hypothesis Test Results")
```

## Conclusion

Our analysis shows that Jeffrey's mean swim time with goggles (approximately `r round(swim_summary$sample_mean, 2)` seconds) is lower than his baseline mean without goggles (`r baseline_mean` seconds). 

The p-value of `r round(swim_summary$p_value, 4)` is less than our significance level of 0.05, suggesting that this improvement is statistically significant. Therefore, we reject the null hypothesis and conclude that wearing goggles does appear to help Jeffrey swim faster.
