---
title: "HW10: Significance Testing & Effect Size"
name: Zoe Bayewitz
---

You previously calculated the mean, margin of error and confidence interval for monthly car thefts in Prince George's County. Your editor now wants to know: it _looks_ like car thefts are decreasing lately, but is that change real, and how big is it? In this assignment, you'll use significance testing and effect size to answer those questions. Load the tidyverse and the data to get started.

### Task 1: Load the tidyverse and the data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
# LOAD THE TIDYVERSE
```

```{r}

# Read the data
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_month.csv") |>
  arrange(month)  # Sort by date

```

### Task 2: Split the data and test for significance (4 points)

We'll split the monthly data into two halves, and then ask whether there's a statistically significant difference between these periods. To do that, we'll use a t-test with our alpha at 0.05. First, fill in the hypotheses, then replace the REPLACE_ME values in the code.

**Null Hypothesis:** From July 2023 to February 2026, there is no statistically significant difference in mean monthly car thefts between the first and second half of the period.

**Alternative Hypothesis:** From July 2023 to February 2026, there is a statistically significant decrease in mean monthly car thefts between the first and second half of the period.

```{r}

# Find the halfway point and add a column placing each month in the first or second halves

midpoint <- ceiling(nrow(car_thefts) / 2)
car_thefts <- car_thefts |>
  mutate(period = if_else(row_number() <= midpoint, "First Half", "Second Half"))

# Perform t-test comparing the two time periods
t_test_periods <- t.test(total ~ period, data = car_thefts)

# Display results
t_test_periods

```

**Answer this question:** What is the p-value? Is it less than 0.05? Can we reject the null hypothesis? What does this tell your editor?

PUT ANSWERS HERE: The p-value is far less than 0.5 (0.0000214); therfore, we can reject the null hypothesis. This tells us that the decline in monthly car thefts is statistically significant and almost definitely not due to random chance.

### Task 3: Visualize the comparison (4 points)

Let's see the two periods side by side with their confidence intervals. Replace the REPLACE_ME values.

```{r}
# Summarize by period
period_summary <- car_thefts |>
  group_by(period) |>
  summarize(
    mean_thefts = mean(total),
    sd_thefts = sd(total),
    months = n(),
    se_thefts = sd_thefts / sqrt(months),
    margin_error = qt(0.975, df = months - 1) * se_thefts,
    ci_lower = mean_thefts - margin_error,
    ci_upper = mean_thefts + margin_error
  )

period_summary |>
ggplot(aes(x = period, y = mean_thefts, fill = period)) +
  geom_col() +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
  geom_text(aes(label = round(mean_thefts, 1)), vjust = -0.5, size = 4) +
  labs(
    title = "Comparing Car Thefts: First Half vs. Second Half",
    subtitle = paste0("p-value = ", round(t_test_periods$p.value, 5)),
    x = "Time Period",
    y = "Average Number of Car Thefts"
  ) +
  scale_fill_manual(values = c("First Half" = "#66c2a5", "Second Half" = "#fc8d62")) +
  theme_minimal() +
  theme(legend.position = "none")

```

**Answer this question:** Look at the error bars (confidence intervals) on the chart. Do they overlap? What does that tell you about whether the difference between the two periods is real?

PUT ANSWERS HERE: The error bars on the chart do not overlap with one another. This tells us that the difference is statistically significant because the ranges of each mean don't overlap and are entirely distinct from one another, showing a dramatic difference.

### Task 4: Calculate the effect size (5 points)

We've established that the decline is statistically significant. But how large is it? Let's calculate Cohen's d to measure the effect size. For a two-sample comparison, we use the pooled standard deviation:

**Pooled SD** = sqrt(((n1 - 1) × sd1² + (n2 - 1) × sd2²) / (n1 + n2 - 2))

**Cohen's d** = (mean1 - mean2) / pooled SD

Replace the REPLACE_ME values. You'll need to pull the values from `period_summary`.

```{r}
# Extract the values from period_summary
first_half_stats <- period_summary |> filter(period == "First Half")
second_half_stats <- period_summary |> filter(period == "Second Half")

# Calculate pooled standard deviation
pooled_sd <- sqrt(
  ((first_half_stats$months - 1) * first_half_stats$sd_thefts^2 +
   (second_half_stats$months - 1) * second_half_stats$sd_thefts^2) /
  (first_half_stats$months + second_half_stats$months - 2)
)

# Calculate Cohen's d
cohens_d <- (first_half_stats$mean_thefts - second_half_stats$mean_thefts) / pooled_sd

# Interpret it
effect_label <- case_when(
  abs(cohens_d) < 0.2 ~ "Negligible",
  abs(cohens_d) < 0.5 ~ "Small",
  abs(cohens_d) < 0.8 ~ "Medium",
  TRUE ~ "Large"
)

tibble(
  `First half mean` = round(first_half_stats$mean_thefts, 1),
  `Second half mean` = round(second_half_stats$mean_thefts, 1),
  `Difference` = round(first_half_stats$mean_thefts - second_half_stats$mean_thefts, 1),
  `Cohen's d` = round(cohens_d, 3),
  `Interpretation` = effect_label
)
```

**Answer this question:** How would you describe this effect size? Compare it to the effect sizes we calculated in class (the response times activity had a small effect, the housing prices activity had a medium effect). What does a large effect size mean for the story you'd write about car thefts?

PUT ANSWERS HERE: This effect size is large, especially when compared to the in-class activities. This shows a dramatically stronger effect than the housing prices and response times. This large effect size tells us that not only have car thefts significantly changed, but that the change was consistent and not random. The gap is large enough to make clear meaningful assumptions about patterns, and that it can be reported on as such.

### Task 5: Putting it all together (7 points)

You now have three pieces of evidence: the confidence intervals from last week, the significance test from this assignment, and the effect size.

1. In Part 1, your editor asked "just tell me if car thefts are going up or down". Now you have significance testing. Write 2-3 sentences telling your editor the answer, using all three pieces of evidence (CI, p-value, and effect size).

PUT ANSWERS HERE: Car thefts are decreasing in Prince George's County, and the data supports this trend. The confidence intervals don't overlap, and a t-test confirms the difference is statistically significant (unlikely to be due to random chance). The Cohen's d tells us that the effect is large, so the decline is not only definite but also substantial.

2. Write a headline and a 2-3 sentence lede for a story about car thefts in Prince George's County. Unlike the response times story from class (where the effect was small and the headline needed to be cautious), this is a case where the data strongly supports a clear conclusion. Your writing should reflect that — but still note the limitations (correlation vs. causation, what else might explain the decline).

PUT ANSWERS HERE:
Car Thefts in Prince George's County Have Plummeted Significantly

Car thefts in Prince George's County have dropped nearly 30% over the past two and a half years, a decline consistent and large enough to rule out random chance. Monthly thefts fell from an average of 439 in the first half of the period to 312 in the second half-- a big enough difference that proves large statistical significance.

When you are finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
