---
title: "Prince George's County Car Theft Spike Analysis"
author: "Your Name"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

## Introduction

This analysis examines the impact of a significant spike in car theft rates in 2023 in Prince George's County, and whether that spike has led to a sustained higher rate of car theft in the county. We'll use Analysis of Variance (ANOVA) to determine if theft rates differ significantly between pre-spike (before 2023), spike (2023), and post-spike (2024-present) periods.

## Data Preparation and Exploration

### Loading Libraries and Data

```{r load-libraries}
# Load required libraries
library(tidyverse)
library(lubridate)
```

```{r load-data}
# Import data
theft_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_month17_25.csv")

# Look at the first few rows
head(theft_data)

# Get a summary of the data
summary(theft_data)
```


### Get the data how we want it and calculate the rates

**Task 1** Replace REPLACE_ME with the variables you need to calculate the theft_rate (2 points)

```{r process-data}
# Calculate theft rate and create time period variable
theft_data <- theft_data |>
  mutate(
    date = as.Date(month),
    year = year(date),
    month_number = month(date),
    theft_rate = (REPLACE_ME / REPLACE_ME) * 100000,
    time_period = case_when(
      year < 2023 ~ "Pre-spike",
      year == 2023 ~ "Spike year",
      year > 2023 ~ "Post-spike"
    )
  )

# Create summary by time period
period_summary <- theft_data |>
  group_by(time_period) |>
  summarize(
    mean_thefts = mean(total),
    mean_rate = mean(theft_rate),
    median_rate = median(theft_rate),
    sd_rate = sd(theft_rate),
    n = n(),
    .groups = "drop"
  )

# Display the summary
knitr::kable(period_summary, digits = 2,
             caption = "Summary of Car Theft Rates by Time Period")
```

**Task 2** Describe the results of calculating the rates for each period. What do they suggest? (3 points)

ANSWER_HERE

## Visualizing the Spike Pattern

### Time Series Plot

```{r time-series}
# Time series plot showing the spike pattern
ggplot(theft_data, aes(x = date, y = theft_rate)) +
  geom_line() +
  geom_smooth(method = "loess", span = 0.2) +
  geom_vline(xintercept = as.Date("2023-01-01"), linetype = "dashed", color = "red") +
  geom_vline(xintercept = as.Date("2024-01-01"), linetype = "dashed", color = "red") +
  annotate("text", x = as.Date("2023-07-01"), y = max(theft_data$theft_rate) * 0.9, 
           label = "Spike Year", color = "red") +
  labs(title = "Monthly Car Theft Rates (2017-2025) with Spike Highlighted",
       x = "Date", y = "Thefts per 100,000 Population") +
  theme_minimal()
```

**Task 3** Describe the output of this chart and then give it a better, more active title that provides the lede (3 points)

ANSWER HERE


Now let's take a look at average theft rates.

### Yearly Average Theft Rates

```{r yearly-means}
# Calculate yearly means
yearly_means <- theft_data |>
  group_by(year) |>
  summarize(
    mean_rate = mean(theft_rate),
    .groups = "drop"
  )

# Plot yearly means
ggplot(yearly_means, aes(x = year, y = mean_rate)) +
  geom_line(size = 1) +
  geom_point(size = 3, color = "blue") +
  geom_point(data = yearly_means |> filter(year == 2023), 
             aes(x = year, y = mean_rate), size = 4, color = "red") +
  labs(title = "Average Annual Car Theft Rates",
       x = "Year", y = "Average Thefts per 100,000 Population") +
  theme_minimal() +
  scale_x_continuous(breaks = 2017:2025)
```

## One-way ANOVA Analysis

### Hypothesis Formulation

For our ANOVA test, we'll compare the mean theft rates between the three time periods. First we need to create our hypotheses.

**Task 4** Fill in the hypotheses (2 points)

- **Null Hypothesis (H0)**: ANSWER HERE

- **Alternative Hypothesis (H1)**: ANSWER HERE

### One-way ANOVA Test

```{r anova-test}
# Perform one-way ANOVA on theft rates by time period
period_anova <- aov(theft_rate ~ time_period, data = theft_data)
summary(period_anova)
```

**Task 5** Describe the result: are there real differences in the average rates between the time periods? Is this result significant? (5 points)

ANSWER HERE

## Compare the differences between specific periods

Run our post-anova Tukey test and examine the results. 

```{r posthoc}
# Tukey's HSD test
tukey_results <- TukeyHSD(period_anova)
print(tukey_results)
```


```{r period-means}
# Visualize time period means with confidence intervals
period_rates <- theft_data |>
  group_by(time_period) |>
  summarize(
    mean_rate = mean(theft_rate),
    se = sd(theft_rate) / sqrt(n()),
    .groups = "drop"
  ) |>
  mutate(
    lower_ci = mean_rate - qt(0.975, n() - 1) * se,
    upper_ci = mean_rate + qt(0.975, n() - 1) * se
  )

# Ensure time period is ordered correctly
period_rates$time_period <- factor(period_rates$time_period, 
                                   levels = c("Pre-spike", "Spike year", "Post-spike"))

ggplot(period_rates, aes(x = time_period, y = mean_rate, fill = time_period)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0.2) +
  labs(title = "Car Theft Rates by Time Period with 95% Confidence Intervals",
       x = "", y = "Mean Theft Rate per 100,000") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Conclusion

This analysis has examined whether there are significant differences in car theft rates between the pre-spike period (before 2023), the spike year (2023), and the post-spike period (after 2023).

**Task 6** Describe the results of the Tukey's test and chart showing confidence intervals. Is there a significant difference between the three periods? In particular, is the difference between the pre-spike and post-spike periods real and significant? Write a headline and lede reporting these findings, focusing on the pre- and post-spike periods. (5 points)

ANSWER HERE
