---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "Zoe Bayewitz"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

In the class activity, you tested a school board member's claim using data from Montgomery County, Pa. — and found a model that didn't hold up. Now you'll run the same analysis using data from Montgomery County, Md. Your job is to evaluate whether the Maryland data tells a different story, and what that means for how you'd report on the claim.

### Loading the Data

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```

### Task 1: Examine the Data First (2 points)

Before you plot anything, look at the data — you know why by now.

```{r}
summary(schools$`FARMS Pct`)
schools |> arrange(desc(`FARMS Pct`)) |> head(10)
```

Are there any data quality issues like the ones you found in the Pennsylvania data? How does the distribution of FARMS percentages compare between the two counties? Note anything that concerns you or gives you confidence in the data.

ANSWER HERE: I don't see as many issues; I'm wondering if they included any other schools and just omitted them in this data (otherwise, what's the point of the "School Type column"?). There are some schools with very high FARMS percentages, and overall lower math scores, but none with 100% FARMS like the Pennsylvania data. This gives me more confidence in the data because there will be fewer outliers.


### Task 2: Create a Scatterplot and Compare (5 points)

Replace the REPLACE_ME value with the correct variable name to create a scatterplot of FARMS percentage vs. math scores for Montgomery County, Md.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations",
       title = "Math Scores vs. FARMS Percentage, Montgomery County MD")
```

Describe what you see. Then compare this scatterplot specifically to the Pennsylvania one: what differences do you notice in the direction, strength, and spread of the relationship? The two counties share a name and are both suburban — what might explain why they look so different?

ANSWER HERE: The two graphs are very different. This graph shows a much stronger negative relationship; the higher a school's FARMS percentage, the lower their math expectations are met. There are still some outlying datapoints here, but none that seem to alter the regression line. I think Montgomery County, MD is going to show a more diverse range of data because while the county is generally considered wealthy, the county is one of the most diverse in the country, in terms of race, religion, and socioeconomic status.

Interesting note: my dad (Mike Bayewitz) was the principal of Broad Acres (poorest school in the county) for a few years and dramatically increased the percentage of students passing state math exams. They talk about this in his bio: https://www.solutiontree.com/michael-d-bayewitz.html


### Task 3: Linear Regression Model (5 points)

Replace the REPLACE_ME value with the correct variable name to run a linear regression model predicting math scores from FARMS percentage.

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)
summary(model)
```

Report the three key statistics: the p-value, the Adjusted R-squared, and the coefficient for FARMS Pct. Then interpret them together — not separately. What does the combination of these three numbers tell you about the relationship between poverty and math scores in Montgomery County, Md.? Is this model worth using? How does it compare to the Pennsylvania model?

ANSWER HERE:
P-value of nearly 0
Adjusted R-squared of 0.6734
Coefficient for FARMS Pct of -0.595

Together, these tell us that poverty and math performance in Montgomery County, MD have a strongly negative relationship. The extremely low p-value tells us that this data is reliable, with little room for chance. The adjusted R means that FARMS percentages explain about 67% of math score variation, which is a very high amount of data covered. And the -0.595 coefficient proves the negative relationship, indicating that for every increase in a student on FARMS, there is a decrease in that school's math performance. This is different from the Pennsylvania data because PA's model proved the data to be unreliable; it also disproved the board member's claim of a negative relationship, which is actually suppoorted by MD's data.


### Task 4: Identifying Schools with the Biggest Residuals (6 points)

Replace the REPLACE_ME values with the correct variable names to calculate predicted values, residuals, and absolute residuals for each school. Then arrange the schools in descending order of absolute residuals.

```{r}
schools <- schools |>
  mutate(
    predicted = predict(model),
    residual = `% Met or Exceeded Math Expectations` - predicted,
    abs_residual = abs(residual)
  ) |>
  arrange(desc(abs_residual))

head(schools |> select(1, `FARMS Pct`, `% Met or Exceeded Math Expectations`, predicted, residual), 10)
```

Identify the school with the largest positive residual (doing much better than the model predicts) and the school with the largest negative residual (doing much worse). For each school, find its zip code and look it up on [CensusReporter.org](https://censusreporter.org).

For each of the two schools, answer:
- What does the demographic and economic profile of the zip code look like?
- Does anything in that profile help explain why the school is over- or under-performing the model?
- What would your first reporting question be to the principal of that school?

ANSWER HERE: Wayside Elementary has the highest positive residual, and Stedwick Elementary has the largest negative residual. Wayside (20854) is 58% White and 22% Asian, with a median household income of $250k. There are 2.6% people below the poverty line. Stedwick (20886) has a median household income of just under $100k, with 12.8% people below the poverty line. The area is 33% Hispanic, 23% White, 23% Black, and 14% Asian. The high economic status and White demographics of 20854 would justify why they are over-performing; people living in this area are less likely to be impoverished or in financial stress, giving them more resources to accomplish educational achievements. Stedwick proves to be more racially diverse, with a much higher poverty rate and lower median income. This could explain why students in that area are under-performing, as their families may not have as many resources or experience financial stress. My first question to the principal of Wayside would be: "What do you think is the biggest reason why your students are able to achieve proportionally high math scores?", and for Stedwick, I'd ask: "What do you think is the #1 resource the school needs to help their students achieve higher test scores?"

### Task 5: Evaluate the Model and Write the Pitch (7 points)

You've now run this analysis on two counties. Return to the school board member's claim: that elementary schools with fewer students on free and reduced-price meals do better on standardized math tests.

Write a short story pitch of 4-6 sentences that addresses all of the following: Does the Maryland data support or contradict the claim? How does it compare to what you found in Pennsylvania? What would you ask the school board member if you called them for comment? And what's the most interesting specific finding from your residuals analysis: which school would you want to visit first, and why?

ANSWER HERE: A school board member in Montgomery County, Pennsylvania claimed that schools with fewer students on FARMS perform better on standardized testing. While the data didn't support his claim, it proved true for a different Montgomery County in Maryland.
Montgomery County, Maryland is one of the most diverse counties in the country, both socioeconomically and racially, and there was a clear correlation between poorer areas, higher FARMS percentages, and low standardized math scores. I would like to speak to the Pennsylvania school board member and ask if he thinks, after looking at Montgomery County MD's results, if he still thinks his has a negative correlation. I'd also speak to the school with the highest and lowest FARMS percentages, and ask why they think their students are performing at the rates that they are.
