---
title: "HW2 Prince George's County Crime"
---

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?

ANSWER HERE: The main categories of incidents seem to be: theft/burglary, accidental crimes, and assaults. Thefts/burglaries can be broken down further into robberies, breaking and entering, and auto thefts. Assaults can include all assaults, homicide, and sex offense. I think that high numbers of theft/burglary/robbery is very significant because it can represent culture, poverty, and safety of any given place-- especially in this case, where the numbers are very high. Different classifications of similar crimes-- for example, vehicle robbery vs. theft from auto-- can make summarizing data difficult because it is hard to draw distinctions between the crimes, and therefore, difficult to know exact numbers and compare/contrast with other crimes. 

### Task 4: Calculating Averages

Calculate the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
pg_crime_averages <- pg_crime |> 
  mutate(total_incidents = Jan + Feb + Mar + Apr + May + Jun +
                      Jul + Aug + Sep + Oct + Nov + Dec) |>
  mutate(average_incidents = (Jan + Feb + Mar + Apr + May + Jun +
                         Jul + Aug + Sep + Oct + Nov + Dec) / 12)
 
```

### Task 5: Describe the Averages

Examine the new dataframe, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?

ANSWER HERE: The categories with the highest average number of incidents per month are: accident, accident with impound, auto stolen, theft, and theft from auto. Residential robbery, sex offense, homicide, and other types of breaking & entering all have low averages. This tells me that the crime in PG County mainly comes from both regular and auto theft. I want to know the distinctions between these similar categories, and what the crimes with "other" include. I also want to know what "accident" and "accident with impound" means.


### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
summer_crime_averages <- pg_crime_averages |>
  mutate(
    june_change = (Jun - average_incidents)/average_incidents * 100, 
    july_change = (Jul - average_incidents)/average_incidents * 100,
    august_change = (Aug - average_incidents)/average_incidents * 100)
```

### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

ANSWER HERE: Certain crimes saw a high percentage increase, like shooting assaults, vehicle robberies, and auto stolen and recovered. These crimes increased in June and July, and then decreased in August. However, others showed percentage decrease, such as theft, accidents, and residential b & e. I think that this may tell a story about vehicle theft increasing in the early summertime; however, otherwise, I didn't find much correlation in crime increases over the summer.

### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

ANSWER HERE: 
Theft and robbery incidents in Prince George’s County are high, with auto-related crimes making up a large share of the total.

Recent crime data shows that theft, both generally and auto-theft, represent a significant portion of offenses in Prince George's County, emphasizing an ongoing public safety concern. These incidents increase, on average, in the early summer months-- June and July-- suggesting a potential correlation between increased auto crime and summer. 


### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
