---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2025.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

ANSWER HERE: Eight employees have a salary of $0, which could be for a variety of reasons. Without their salaries, we can't assure that the information will be entirely accurate. 

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv")

wh_salaries |>
summarize(
  mean_salary = mean(SALARY),
  median_salary = median(SALARY)
)
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |>
summarize(
  mean_salary = mean(SALARY),
  median_salary = median(SALARY),
  min_salary = min(SALARY),
  max_salary = max(SALARY)
)
  
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

ANSWER HERE: The mean and median salaries tell me that the average white house employee makes over $100k annually. Since some of the data was wrong, as there were multiple salaries with a value of 0, we can assume that the mean and median would probably be higher. The maximum salaries tell us that advisors typically make the most money. The outliers of the data set are the 8 employees with undisclosed salaries.

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

ANSWER HERE: My story idea is investigating the salaries of the 8 employees (all who have jobs closely working with the President) and seeing why their salaries are undisclosed. To develop the story, I would attempt to interview current and/or former White House employees, and/or people close to them like family members. I would also look at datasets from past years and see if this is something typically done in White House administrations. I would specifically look at the salaries of these 8 individuals in the past, as well as the salaries of those with their job titles (if the people are different), and compare to the current data.
