---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities |>
  mutate(
    violent_rate_2019 = (violent2019/pop2019)*10000,
    property_rate_2019 = (property2019/pop2019)*10000,
    violent_rate_2020  = (violent2020/pop2020)*10000,
    property_rate_2020 = (property2020/pop2020)*10000
   )
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

ANSWER HERE: 
Cities like Baltimore and Ocean City have seen high crime rates, both with violent and property crime. These rates remain high in these cities, which suggests that crime is concentrated in certain areas rather than at random. Some cities, like Oxford and Queen Anne, saw a decrease in property crime, while cities like Edmonston and Laytonsville saw big increases in violent crime. Most of the cities with high crime rates are heavily-populated, although Baltimore's concentration of crime remains high. Baltimore saw slight decreases from 2019 to 2020 in both types of crimes, suggesting that COVID led to a decrease of violent and property crimes there.


### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates |>
  mutate(
    violent_rate_pct_change  = ((violent_rate_2020-violent_rate_2019)/violent_rate_2019)*100,
    property_rate_pct_change = ((property_rate_2020-property_rate_2019)/property_rate_2019)*100
  )
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

ANSWER HERE: One interesting thing I notice is that multiple cities in Prince George's county saw a high percentage increase in violent crime, suggesting increasing violence in the county. Montgomery and Prince George's counties, two highly populated areas, saw increases in property crime as well. Several small cities in Maryland saw increased property crime rates, but since those areas are low-populated, a significant jump in crime could be a difference of just a few people. Overall, I think the most interesting data is the violent crime rate increase in Prince Gerorge's county.

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

ANSWER HERE: Violent crime rose across Prince George’s County from 2019 to 2020, even as much of daily life shut down during the first year of the COVID-19 pandemic.

Cities like Edmonston, Cottage City, Glenarden and Mount Rainier saw violent crime rates increase by over 60% in a single year. Despite widespread lockdowns and stay-at-home orders in 2020, the data show that violent crime did not slow in many of the county’s smaller cities, underscoring how public safety challenges persisted during an unprecedented public health crisis.

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
