---
title: "JOUR405: Statistics for Journalists"
subtitle: "Midterm Exam"
name: "Zoe Bayewitz"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# JOUR405: Statistics for Journalists
## Midterm Exam

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences.

## Load the tidyverse

```{r}
library(tidyverse)
library(ggplot2)
```

---

## Part 1: Restaurant Health Inspections (25 points)

The first dataset contains restaurant health inspection scores from Montgomery County. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv`

The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible, the compliance score, and the grade.

### Task 1: Read the data (2 points)

```{r}
inspections <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")
```

### Task 2: Calculate the mean, standard deviation, minimum, and maximum of compliance scores. Save the results in a new dataframe called `compliance_stats`. (2 points for each)

```{r}
compliance_stats <- inspections |> 
  summarize(mean_compliance = mean(compliance_score),
            sd_compliance = sd(compliance_score),
            min_compliance = min(compliance_score),
            max_compliance = max(compliance_score))

compliance_stats
```

### Task 3: Create a histogram of the compliance scores with a vertical line showing the mean. Include a descriptive title and label the x and y axes using `labs()`. (10 points)

```{r}
  ggplot(inspections) + 
  geom_histogram(aes(x = compliance_score), binwidth = 2) +
  geom_vline(aes(xintercept = mean(compliance_score)), color = "red", linetype = "dashed", size = 1) +

  labs(
    title = "Compliance Scores of Montgomery County Health Inspections" ,
    x = "Compliance Score" ,
    y = "Frequency")
  
```

### Task 4: Write 2–3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What is newsworthy, and what would your next reporting step be? (5 points)

**Answer:**

The histogram tells us that most compliance scores are high, indicating that most restaurants are within the necessary score, with some restaurants getting lower scores. The standard deviation tells us that most restaurants come within a few points of the average high compliance score, so the majority of all restaurants are getting a good enough score. Something newsworthy would be finding the establishments with the lowest and highest scores, interviewing them both, and comparing/contrasting their safety methods. I would look at the restaurants with the lowest scores, find a few restaurants with high scores, and interview them both about their health practices.
---

## Part 2: Daily Transit Ridership — Z-Scores and Sampling (30 points)

Your editor wants you to identify unusual days in WMATA's daily ridership data that might be worth investigating as story leads. The dataset contains daily bus and rail totals. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv`

### Task 5: Load the WMATA ridership data. (2 points)

```{r}
wmata_daily <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")
```

### Task 6: Using `summarize()`, calculate the mean and standard deviation for both bus and rail ridership. Save the results in a dataframe called `wmata_stats`. (5 points)

```{r}
wmata_stats <- wmata_daily|> 
  summarize(
    mean_bus = mean(bus),
    sd_bus = sd(bus),
    mean_rail = mean(rail),
    sd_rail = sd(rail))

wmata_stats

```

### Task 7: Using `mutate()`, add two new columns to `wmata_daily`: `rail_zscore` (the z-score for rail ridership) and `bus_zscore` (the z-score for bus ridership). Save the result as `wmata_zscores`. (8 points)

```{r}
wmata_zscores <- wmata_daily |>
  mutate(
    rail_zscore = (total - mean(rail)) / sd(rail), 
    bus_zscore = (total - mean(bus)) / sd(bus)
  )

wmata_zscores
```

### Task 8: Filter `wmata_zscores` to show only days where the absolute value of `rail_zscore` is greater than 2 (unusually high or low rail ridership). Arrange by the absolute value of the z-score, with the most extreme first. How many unusual days are there? (8 points)
There are 301 days with unusual rail z-scores.

```{r}
unusual_days <- wmata_zscores |>
  filter(abs(rail_zscore) > 2) |>
  arrange(desc(abs(rail_zscore)))

cat("Number of days with unusual rail z-scores:", nrow(unusual_days), 
    "out of", nrow(wmata_zscores), "total days (", 
    round(nrow(unusual_days)/nrow(wmata_zscores)*100, 2), "%)")
```

### Task 9: Using `set.seed()`, take a random sample of 91 days from `wmata_daily` (roughly 25% of the data). Calculate the mean and standard deviation for bus and rail ridership in the sample. How do the sample statistics compare to the population statistics you calculated in Task 6? Write 1–2 sentences comparing the values and explaining what this suggests about random sampling. (4 points)

Compared to the values calculated in Task 6, this sample's mean slightly decreased and standard deviation slightly increased for both bus and rail. This suggests that while taking a sample will often get you similar results to the total data, results may be altered, and the more data you take into account, the more accurate the calculations will be.

```{r}
set.seed(89)

sample_91 <- wmata_daily |>
  slice_sample(n = 91)

sample_91 |>
  summarize(
    mean_bus = mean(bus),
    sd_bus = sd(bus),
    mean_rail = mean(rail),
    sd_rail = sd(rail)
  )
```

### Task 10: Write 2–3 sentences: What do the unusual days in your z-score analysis suggest as story leads? (3 points)

Most of the unusually high z-scores are on Tuesdays, Wednesdays, and Thursdays. This suggests that people are traveling to work more on those days, and less on Mondays and Fridays. This could lead to a story about office's transitions to hybrid/remote work models post-Covid, allowing workers to work from home at the beginning and end of the week.
---

## Part 3: Maryland Car Theft Rates (25 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to find out which counties have the highest rates. The dataset contains car theft counts for Maryland counties in 2023 along with population figures. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv`

### Task 11: Load the car thefts data. (2 points)

```{r}
md_car_thefts <- read_csv ("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")
```

### Task 12: Using `mutate()`, add a column called `theft_rate` that calculates the car theft rate per 10,000 residents for each county. Save the result back to `md_car_thefts`. In 1–2 sentences, explain why a per-capita rate is more useful than raw counts for this comparison. (8 points)

A per-capita rate is more useful than raw counts for this comparison because every county in Maryland has a different population size, so comparing the numbers plainly would unfairly show higher thefts in higher populated counties, and vice versa. By using a rate to compare all of the counties, we can analyze their crime in comparison to each other rather than by individual populations.

```{r}
md_car_thefts <- md_car_thefts |> 
  mutate(theft_rate = (thefts_2023/population) * 10000)
```

### Task 13: Calculate the median car theft rate and the total number of car thefts statewide. Show the counties that have rates above the median. (7 points)

```{r}
median_rate <- md_car_thefts |>
  summarize(median_rate = median(theft_rate))

total_thefts <- md_car_thefts |>
  summarize(total_thefts = sum(thefts_2023))

high_median <- md_car_thefts |>
  filter (theft_rate > median(theft_rate)) |>
  select (county, theft_rate)

high_median
```

### Task 14: Write 2–3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What is the lede of a story based on your findings? (8 points)

These calculations reveal that the more populated counties in Maryland (Baltimore, Montgomery, PG, Howard) have higher theft rates than lower populated counties, and Baltimore City has a much higher theft rate than any other county. Based on these findings, I would write a story about car thefts in Baltimore, comparing car thefts in the city versus in Baltimore County and analyzing why there is such a high concentration of theft within the city.

---

## Part 4: Data Analysis Scenario (20 points)

A city parks department releases data showing that its 12 community swimming pools average 187 visitors per day during the summer season, with a standard deviation of 42 visitors. Your editor suspects that a handful of pools are dramatically underperforming and wants a story. You also learn that the department's figures are based on a random sample of 60 days per pool rather than a complete count.

Write 3–4 sentences for each of the following questions:

### Question A: Given that the mean is 187 and standard deviation is 42, which statistical measures would you prioritize using and why? (7 points)

I would prioritize using the mean, standard deviation, and z-score to analyze the data. Since the mean is high and there's a large standard deviation, these measures would help easily identify the pools that are underperforming. By analyzing pools that fall below the mean and the standard deviation, we can tell which pool's have unusually low attendance and are underperforming.

### Question B: How would you use z-scores to identify which pools are genuinely underperforming, and why is that more useful than simply ranking pools by raw attendance? (7 points)

I would use the z-score to compare the pool's attendance to the overall mean and standard deviation. Pools that are genuinely underperforming would likely have a highly negative z-score, indicating that their attendance is lower than what is in the typical range. This approach is more useful than ranking pools by raw attendance because it compares the attendance to the overall distribution (a pool may have low attendance but still be within normal range), and will show which pools have a highly unusual low attendance rate.

### Question C: The department collected its attendance figures using a random sample of 60 days per pool rather than tracking every day. What questions would you ask about the sample before relying on these figures for a story? What concerns would you raise in your story? Use statistical terminology where appropriate. (6 points)

Before relying on these figures for a story, I would ask how the random sample was conducted. For example, which 60 days were sampled-- was it all in the summer, did it include holidays, and did it include both weekdays and weekends? Not knowing these days would raise concern, as the pool is most attended during summer weekends and holidays, and using data from days where the pool was closed or where there was bad weather would increase the likelihood of the pool's z-score indicating underperformance. I would also ask if 60 days is a large enough sample to represent the entirety of summer attendance, and what the range of dates are-- does this range accurately represent attendance from the pools opening (usually Memorial Day) to closing (usually Labor Day), or did it just do the months of June, July, and August? Why wasn't a bigger sample conducted?
---

### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your choices

Good luck!
