---
title: "Final Project"
output: html_notebook
---


HYPOTHESIS: Revenue earned from streaming services has not fully compensated for the music industry's decline in physical music sales, when adjusted for inflation.

Code I could run: 
- adjusting inflation from monthly to yearly (partially done-- just need to calculate averages)
  - compare this data to physical and streaming sales
- linear regression chart: find the lines of streaming and physical revenue over time and graph them using a ggplot
- (more math) calculate yearly percent changes of physical and streaming sales (gives us more specific indicators of when streaming didn't compensate for physical)
- measure correlation coefficient of physical's decline and streaming's rise 
- calculate the dollar amount between what physical earned before and what streaming earns now, including inflation

```{r}
library(tidyverse)
```

```{r}
music_revenue <- read_csv("/Users/zoebayewitz/Documents/GitHub/jour405_spring26/project/Recorded_Music_Revenue.csv")
```
Historical data of yearly revenue recorded by RIAA of music sales from 1973-2025. Filtered by four formats:  Physical, Digital Downloads, Streaming, and Synchronization. Physical versus Streaming will be my focus. The data is originally in wide format, so it needs to be switched into long format. I'll focus on the physical and streaming rows (use Derek code). I'll also need to clean the data because numbers are in revenue format, and the column names aren't cleaned.

```{r}
music_revenue_long <- music_revenue |>
  pivot_longer(
    cols = -Format2,
    names_to = "year",
    values_to = "revenue"
  ) |>
  mutate(
    year = as.integer(year),
    revenue = parse_number(revenue)
  ) |>
  rename(format = Format2)
```


```{r}
inflation <- read_csv("/Users/zoebayewitz/Documents/GitHub/jour405_spring26/project/cpi_inflation.csv")
```
Tracks monthly inflation rates from 2015-2022, to be compared against streaming sales. I need to add data going back to 1973. I'll then need to reformat the monthly data by turning it into an annual figure (finding averages). Other than that, I just need the CPIAUCSL column to measure it against the revenue; .

TRANSFORM MONTHLY INTO ANNUAL:
```{r}
inflation_annual <- inflation |>
  rename(date = observation_date, cpi = CPIAUCSL) |> 
  mutate(
    date=as.Date(date),
    year=as.integer(format(date, "%Y"))
  )

```

